Generating Correlated Data for Omics Simulation

### 2025-06-02 manuscript/paper.qmd
Incorporate metabolomics dataset

@@

We highlight that our methods are highly efficient and easily scale to the large feature counts typical of omics data sets.
While runtime of simulators is typically not critical, we hypothesize that it introduces an inconvenience that has slowed the adoption of dependence in omics simulation and could be particularly cumbersome whenever large numbers of simulated datasets are needed.
Of note is that alternative methods are typically run with only a subset of genes chosen, for example, 500-200 genes for scDesign2 [@Sun2021], 1000 genes for scDesign3 [@scDesign3], and 621 for SeqNet [@SeqNet].
In contrast, our methods can operate on the entire space of genes, avoiding an arbitrary choice of cutoff, though it should be noted that correlation is most meaningful in high-expressed, high variance genes so subsetting genes is also a reasonable approach.

@@

While we describe one special form of $\Sigma_{sim}$ where generation is highly efficient, there exist many other possiblecovariance matrix forms.
Block diagonal covariance matrices are one such example which can efficiently represent even high-dimensional data [@Devijver02012018] and lend themselves to efficient simulation.
Sparse matrix methods [@Bien2011-nx, @Bickel_sparse, @Cai_adaptive] yield exactly zero correlation between some variables, but it is less clear how to simulate data from these efficiently.
Iterative sampling methods [@Krylov_subspaces] are one approach that could exploit the sparse nature and may even extend to sparse inverse covariance methods [@sparse_inverse_covariance].

@@

Source code for all simulations and figures in this paper is available at [github.com/itmat/dependent_sim_paper/](https://github.com/itmat/dependent_sim_paper/).
Source code for the `dependentsimr` package is available at [github.com/tgbrooks/dependent_sim](https://github.com/tgbrooks/dependent_sim/).
All data used is available from the Gene Expression Omnibus (GEO) with accession numbers GSE151923, GSE81142, and GSE151565.

>>>>

We further demonstrate our methods on a different data type, metabolomics, where we use normal marginal distributions and with far fewer features measured (42), see Supplemental Figure 3.

@@

We highlight that our methods are highly efficient and easily scale to the large feature counts typical of omics data sets.
While runtime of simulators is typically not critical, we hypothesize that it introduces an inconvenience that has slowed the adoption of dependence in omics simulation and could be particularly cumbersome whenever large numbers of simulated datasets are needed.
Moreover, high memory use of the alternatives could be an even larger barrier than run-time.
Of note is that alternative methods are typically run with only a subset of genes chosen, for example, 500-200 genes for scDesign2 [@Sun2021], 1000 genes for scDesign3 [@scDesign3], and 621 for SeqNet [@SeqNet], presumably due to the resource requirements.
In contrast, our methods can operate on the entire space of genes, avoiding an arbitrary choice of cutoff, though it should be noted that correlation is most meaningful in high-expressed, high variance genes and so subsetting genes is a reasonable approach.

@@

While we describe one special form of $\Sigma_{sim}$ where generation is highly efficient, there exist many other possible covariance matrix forms.
Block diagonal covariance matrices are one such example which can efficiently represent even high-dimensional data [@Devijver02012018] and lend themselves to efficient simulation.
Sparse matrix methods [@Bien2011-nx, @Bickel_sparse, @Cai_adaptive] yield exactly zero correlation between some variables, but it is less clear how to simulate data from these efficiently.
Iterative sampling methods [@Krylov_subspaces] are one approach that could exploit the sparse nature and may even extend to sparse inverse covariance methods [@sparse_inverse_covariance].

@@

Source code for all simulations and figures in this paper is available at [github.com/itmat/dependent_sim_paper/](https://github.com/itmat/dependent_sim_paper/).
Source code for the `dependentsimr` package is available at [github.com/tgbrooks/dependent_sim](https://github.com/tgbrooks/dependent_sim/).
All sequencing data used is available from the Gene Expression Omnibus (GEO) with accession numbers GSE151923, GSE81142, and GSE151565.
Metabolomics data was obtained from MetaboAnalyst (https://www.metaboanalyst.ca/MetaboAnalyst/upload/RocUploadView.xhtml).
